---
title: "Global NDVI & Population HIA"
author: "Michael D Garber"
date: "`r Sys.Date()`"
output:
  bookdown::html_document2:
    toc: true
    toc_float: true
    toc_depth: 4
---

# Summary

## Motivation

Ideal levels of green space may differ depending on density. At the global scale, there is a need to describe greenness appropriate for its region and population-density context. Such knowledge would identify feasible target areas for improved greening at the local level allowing to estimate the health benefits of such scenarios.

## Objective

Estimate the health impacts of urban green scenarios based on population density-stratified measures of greenness in cities around the world.

## Methods

### Data sources

-   Green space (NDVI)
    -   Description: As described by Chinese colleagues: NDVI data: MODIS NDVI 16-day composite product at a 1 km resolution (MOD13A2 Version 6), we calculated the annual maximum value of NDVI for each pixel every year on the Google Earth Engine (GEE) platform.
    -   Year = 2019
    -   Spatial resolution = 1km
    -   Coverage: global
-   Population
    -   Landscan population data
        -   I downloaded from <https://landscan.ornl.gov>
        -   Population data are formatted in terms of RGB values:
            -   1 - 5: light yellow rgb(255,255,190)
            -   6 - 25: medium yellow rgb(255,255,115)
            -   26 - 50: yellow rgb(255,255,0)
            -   51 - 100: orange rgb(255,170,0)
            -   101 - 500: orange-red rgb(255,102,0)
            -   501 - 2500: red rgb(255,0,0)
            -   2501 - 5000: dark red rgb(204,0,0)
            -   5001 - 185000: maroon rgb(115,0,0)
        -   spatial resolution = 1 km
        -   year = 2019
        -   coverage = global
-   Estimates of mortality rates
    -   United Nations number of deaths (adults 20+) and population size (adults 20+) for every country
        -   <https://population.un.org/wpp/Download/Standard/Mortality/>
    -   Following approach used in global paper with David and Chinese colleagues
    -   I down-weighted the Landscan data by the ratio of 20+ to total population in every country per UN
-   Global urban boundaries
    -   article: <https://iopscience.iop.org/article/10.1088/1748-9326/ab9be3#erlab9be3s>
    -   data downloaded from <http://data.ess.tsinghua.edu.cn/gub.html>
-   Biomes / ecoregions
    -   Ecoregions 2017 is from <https://ecoregions.appspot.com>
    -   found via <https://developers.google.com/earth-engine/datasets/catalog/RESOLVE_ECOREGIONS_2017#description>

### Approach

-   Stratify by biome, then by city, then by Landscan population category and then:

-   Measure tertiles of NDVI in each biome-city-pop-group stratum.

-   Scenario:

    -   Set the NDVI of pixels in the bottom two tertiles to the NDVI value of the 83rd percentile (median of top tertile). In other words, only intervene upon pixels in the bottom two tertiles for that population category for that city for that biome. Note there are a few cities where biome varies within city.

    -   The idea is that this would be a realistic intervention given the target NDVI is relative to the biome, city, and population density category.

-   Conduct HIA using mean of Landscan populatoin values in that category (also plan to use min/max for uncertainty analyses).

## Status and next steps

HIA analysis complete for continental USA (48 states+DC) following those steps.

Working on expanding to global.

Discussion question: consider restricting to cities above a certain population?

# Figures and tables

## Depiction of data sources 
### Map of biomes (Continental USA)

```{r, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
library(mapview)
library(tidyverse)
library(sf)
library(here)
#from global-ndvi-pop/scripts/read-ecoregions-biomes.R
setwd(here("data-processed"))
load("biomes_14_usa_48.RData")
load("biomes_14_usa_48_simplified.RData")
#use smaller file for vis
biomes_14_usa_48_simplified %>% 
  mapview(
    zcol = "BIOME_NAME",
    col.regions = viridis::turbo(
      n=n_distinct(biomes_14_usa_48_simplified$BIOME_NAME)
      ),
    layer.name ="BIOME_NAME"
    )
```

### Map of Landscan population categories (Colorado)

The map visualizes values categories coded 1-8 for easier visualization. The corresponding population categories appear in the table below.

```{r, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
library(raster)
library(tidyterra)
setwd(here("data-processed"))
ls_2019_co_wrangle = terra::rast("ls_2019_co_wrangle.tif")  
#names(ls_2019_co_wrangle)
ls_2019_co_wrangle %>% 
  as_tibble() %>% 
  group_by(pop_cat_1_8) %>% 
  summarise(
    pop_cat_min_val = mean(pop_cat_min_val, na.rm=TRUE),
    pop_cat_max_val = mean(pop_cat_max_val, na.rm=TRUE),
    pop_cat_mean_val = mean(pop_cat_mean_val, na.rm=TRUE)
    )

mapviewOptions(
  basemaps = c("CartoDB.DarkMatter", "CartoDB.Positron")
)
ls_2019_co_wrangle$pop_cat_max_fac %>%
  raster::raster() %>% 
  mapview(
    layer.name = "pop_cat_1_8",
    zcol = "pop_cat_1_8")
```

## Map of global urban boundaries (Colorado)

Data are large, so only mapping Colorado. Visualize the area (square kilometers) of urban boundaries.

```{r, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
setwd(here("data-processed"))
load("gub_colorado_simplified.RData")
mapviewOptions(
  basemaps = c("CartoDB.Positron","CartoDB.DarkMatter")
)
gub_colorado_simplified %>%
  mapview(
    zcol = "area_km2",
    layer.name ="area_km2")
```

## Results (Continental USA)
### Results by biome

```{r, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
setwd(here("data-processed"))
load("pop_ndvi_gub_biome_usa_48_tib.RData")
source(here("scripts", "analysis-functions.R")) #for hia_summarise() function
t_biome_usa = pop_ndvi_gub_biome_usa_48_tib %>% 
  group_by(biome_name_imp) %>% 
  hia_summarise() %>% 
  dplyr::select(
    starts_with("biome"), starts_with("pop"),
    starts_with("ndvi"),starts_with("deaths") )
t_biome_usa %>% 
  knitr::kable(
    caption = "HIA results by biome - Continental USA",
    #remove underscores in column names. easy way to make it more legible.
    #https://bookdown.org/yihui/rmarkdown-cookbook/kable.html
    col.names = gsub("[_]", " ", names(t_biome_usa)),#remove underscores
    format.args = list(big.mark = ",", scientific = FALSE),
    digits = c(0,0,2,2,2,0,0,3)
  )

```
### Results by city
Cities above 1,000,000 people (per Landscan) sorted ascending by deaths prevented per 1k pop (top 10)
```{r, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
setwd(here("data-processed"))
load("lookup_gub_city_name.RData")
hia_summary_usa_48_gub = pop_ndvi_gub_biome_usa_48_tib %>% 
  group_by(ORIG_FID) %>% 
  hia_summarise() %>% 
  arrange(desc(deaths_prevented_per_1k_pop_mean)) %>% 
  slice(1:100) %>% 
  left_join(lookup_gub_city_name, by = "ORIG_FID") %>% 
  dplyr::select(ORIG_FID, city_name, everything())

hia_summary_usa_48_gub
```

### Map: Deaths prevented per 1,000 population by city
Results are presented at the level of the global urban boundary. Following the methods described above, they are first stratified by biome and are thus relative to biome within urban boundary if biome varies within urban boundary.

```{r, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
setwd(here("data-processed"))
load("gub_usa_48_hia_simplified.RData")
library(viridis)
mapviewOptions(
  basemaps = c("CartoDB.DarkMatter", "CartoDB.Positron")
)
gub_usa_48_hia_simplified %>% 
  mapview(
    lwd=.1,
    col.regions = viridis_pal(option = "plasma"),
    layer.name = "Deaths prevented per 1k pop, mean",
    zcol = "deaths_prevented_per_1k_pop_mean")
```


## Results (Global)
As of February 20th, we exclude global urban areas with a population below 1,000 (minimum Landscan bound) and below 5 square kilometers.
### How many cities included?
```{r, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
setwd(here("data-processed"))
load("pop_ndvi_gub_biome_tib.RData")
hia_summary_gub= pop_ndvi_gub_biome_tib %>% 
  group_by(ORIG_FID) %>% 
  hia_summarise() 
hia_summary_gub %>% 
  nrow()
```

### Popluation distribution of cities included
Based on summarized Landscan 2019 data (mean of categories)

```{r, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
hia_summary_gub %>% 
  dplyr::select(pop_cat_mean_val_scaled) %>% 
  summary(pop_cat_mean_val_scaled)
```
### Area distribution of cities included
```{r, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
setwd(here("data-processed"))
load("gub_nogeo.RData")
gub_nogeo %>% 
  dplyr::select(area_km2) %>% 
  summary(area_km2)
```


### Results by biome
```{r, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
setwd(here("data-processed"))
load("pop_ndvi_gub_biome_tib.RData")
t_biome_global = pop_ndvi_gub_biome_tib %>% 
  group_by(biome_name_imp) %>% 
  hia_summarise() %>% 
  dplyr::select(
    starts_with("biom"), starts_with("pop"),
                starts_with("ndvi"),starts_with("deaths") )
t_biome_global %>% 
  knitr::kable(
    caption = "HIA results by biome - global",
    col.names = gsub("[_]", " ", names(t_biome_global)),#remove underscores
    format.args = list(big.mark = ",", scientific = FALSE),
    digits = c(0,0,2,2,2,0,0,3)
  )

```


### Results by city
Cities above 1,000,000 people (per Landscan) sorted ascending by deaths prevented per 1k pop (top 20)
```{r, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
setwd(here("data-processed"))
source(here("scripts","generate-boundaries-states-countries.R"))#
load("lookup_gub_geoname_id.RData")#gub geoname id lookup created ~read-gub.R

load("hia_summary_gub_1mil_plus.RData")#analysis
hia_summary_gub_1mil_plus_4_vis = hia_summary_gub_1mil_plus %>% 
  left_join(lookup_gub_geoname_id, by = "ORIG_FID") %>% #lookup city id
  left_join(lookup_geoname_id_city_name, by = "geoname_id") %>% #lookup city name
  left_join(lookup_geoname_id_country_name, by = "geoname_id") %>% #country name
  dplyr::select(
    starts_with("city_name"),
    starts_with("country_name"), 
    starts_with("pop"),
    starts_with("ndvi"),starts_with("deaths")
  ) 
hia_summary_gub_1mil_plus_4_vis %>% 
  arrange(desc(deaths_prevented_per_1k_pop_mean)) %>% 
  slice(1:20) %>% 
  knitr::kable(
    caption = "HIA results by city - top 20 deaths prevented per pop.",
    col.names = gsub("[_]", " ", names(hia_summary_gub_1mil_plus_4_vis)),#remove underscores
#  digits = c(0,0,0,2,2,2,0,0,3)
    format.args = list(big.mark = ",", scientific = FALSE),
  )
```

### Map: Deaths prevented per 1,000 population by city (Top 100) among cities with 1 million+ population

```{r, eval=TRUE, echo=FALSE, warning=FALSE, message=FALSE}
library(viridis)
setwd(here("data-processed"))
load("gub.RData")
load("hia_summary_gub_1mil_plus.RData")#analysis
mapviewOptions(
  basemaps = c("CartoDB.DarkMatter", "CartoDB.Positron")
)

hia_summary_gub_1mil_plus %>% 
  arrange(desc(deaths_prevented_per_1k_pop_mean)) %>% 
  slice(1:100) %>% 
  left_join(gub, by = "ORIG_FID") %>% 
  left_join(lookup_gub_city_name, by = "ORIG_FID") %>% 
  st_as_sf() %>% 
  st_centroid() %>% #convert to point for better vis
  mapview(
    lwd=.1,
    col.regions = viridis_pal(option = "plasma"),
    layer.name = "Deaths prevented per 1k pop, mean",
    zcol = "deaths_prevented_per_1k_pop_mean")
```



